---
title: "Vignette for the functions of the athi package"
author: Detlef Groth, University of Potsdam, Germany
date: 2023-08-01
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Vignette for the functions of the athi package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
include-before: |
    <style> body { max-width: 1000px; font-family: Candara, sans-serif;} </style>
    <center>[athi functions](athi-functions.html) - [athi analysis](athi-data-analysis.html) </center>
    
---

```{r echo=FALSE,results="hide"}
library(knitr)
```

## Introduction

Below you find example code for the package for most of the  functions  which
should illustrate on how to use them.

You first should load the package using the standard  `library`  function. All
functions are provided  within an  environment  `athi` to see what methods are
available you can use the `ls` command:

```{r}
.libPaths(c("~/R",.libPaths()))
library(athi)
ls(athi)
```

Alternatively all functions have copies using the athi-underline  prefix which
you can as well view like this:

```{r}
ls(pattern="athi_")
```

The  advantage of the  environment  athi is, that you can easily extend it by
adding  your own data and  functions  to it. It is then  possible to save this
file  as RDS and to give it to  someone  else or as a  supplement  for a paper
without the need to create a full package.

Here an example:

```{r}
athi$hello = function () { print("Hello World") }
data(iris)
athi$data_iris = iris
ls(athi)
```

You      can       save       this       extended       environment       with
`saveRDS(athi,file="athi-extended.RDS")`  and you can  load  this  environment
then on other computer using the command `athi=readRDS("athi-extened.RDS")`


## Functions 

### athi$assoc_plot - contingency table plot with effect size measure

Create an association plot with Pearson residual coloring and legend.

This function updates the standard assocplot function from the graphics package 
with the ability to display residual colors. In blue and red are shown groups with 
residuals  above +4 or below -4 in light colors are shown residuals  between 2
and 4 for positive and -4 and -2 for negative  residuals.  At the bottom there
is the residual legend shown as well as the P-value for the Chisq-test and the
effect size measure Cohen's d.

Usage: *athi_assoc_plot(x,legend=TRUE,shade=TRUE,...)*

```{r fig=TRUE,fig.width=5,fig.height=5}
par(mai=c(0.7,0.7,0.3,0.1))
x <- margin.table(HairEyeColor, c(1, 2))
athi$assoc_plot(x)
```

### athi$bootstrap - resampling for empirical p-values or quantiles

Perform a resampling for the given data set and function. 

The function allows you to perform a resampling method without replacement to perform
a boostrap analysis for instance to cmpute a p-value or a confidence interval.

Usage: *athi_bootstrap(x,FUN=NULL,n=1000,...)*

```{r}
rn=rnorm(100,mean=10,sd=2)
t.test(rn)$conf.int
vals=athi$bootstrap(rn,FUN=mean)
summary(vals)
quantile(vals,c(0.025,0.975)) # 95% CI is very close
### confidence interval for spearman correlation
cor.test(swiss[,1],swiss[,2],method="spearman")
vals=athi$bootstrap(swiss[,c(1,2)],
      FUN=function(x) cor(x[,1],x[,2],method="spearman"))
summary(vals)
### insignificant as it crosses zero
quantile(vals,c(0.025,0.975)) # 95% CI shows insignifcant
```

### athi$box_plot - boxplot with effect size measure

Visualize a mean comparison with boxplot and main statistics.

This function is plotting the standard boxplot for a numerical data against a factor variabl and adds on top the 
the main statistics, like the Cohen's d, or the Eta-squared value, the significance level and for t-tests the  
confidence interval as well.

Usage: *athi_box_plot(x,y=NULL,data=NULL,col='grey80',grid=TRUE,main=NULL,...)*

```{r fig.width=5,fig.height=4}
data(iris)
athi$box_plot(Sepal.Length ~ Species, data=iris,col=2:4)
```

### athi$cdist - correlation distance measure

Correlation  distances  between  two  vectors or for all  variable  pairs in a
matrix.

Usage: *athi_cdist(x,method="pearson",type="abs")*

```{r}
### Euclidean distance
round(dist(t(scale(iris[,1:4]))),2)
### Correlation distance
round(athi$cdist(t(iris[,1:4])),2)
round(athi$cdist(t(iris[,1:4]),method="spearman",type="square"),2)
```

### athi$chr2ord - characters or factor to number conversion

Map characters or factors to numbers.

Create ordinal numerical variables out of character or factor variables based on a given mapping.

Usage: *athi_chr2ord(x,map)*

```{r}
status=c("never","rare","often","always")
x=sample(status,100,replace=TRUE)
x=c(NA,x,NA)
table(x,useNA='ifany')
map=c(never=0, rare=1, often=2,always=3)
table(athi$chr2ord(x,map),useNA='ifany')
```

### athi$cihist - confidence intervals for density lines

This is a utility function to draw a histogram with density 
lines and the confidence level for that density line.

Usage: `athi_cihist(x, conf.level=0.95, legend=TRUE, xlab="Value", ylab="Density", ...)`

```{r fig.width=6,fig.height=4}
set.seed(123)
data <- rnorm(1000, mean = 0, sd = 1)
athi$cihist(data,main="Histogram density line with 95% confidence level")
```


### athi$cohensD - effect size for comparing two means

Effect sizes measure for difference between two means.

Calculate the effect size for the difference between two means, divided by the pooled standard deviation.

The function cohensD calculates the effect size for the difference between two means.
Due to Cohen's rule of thumb values of around 0.2 are considered to stand 
for small effects, values of around 0.5 represent medium effects and values of around 0.8 
and larger represent large effects. 

Please note that these rules of thumb are not useful for highly dependent 
outcome variables (death for instance) these rules might not be useful and as 
well lower values might be of practical relevance.

Usage: *athi_cohensD(x,g,paired=FALSE)*
 
```{r}
cohensD=athi$cohensD
set.seed(125)
data(sleep)
with(sleep,cohensD(extra,group))
x1=rnorm(100,mean=20,sd=1)
x2=rnorm(100,mean=22,sd=1)
g1=rep('A',100)
g2=rep('B',100)
### difference should be around 2SD
cohensD(c(x1,x2),as.factor(c(g1,g2)))
### biseriell correlation coefficient as alternative
### value is as well large
cor(c(x1,x2),as.numeric(as.factor(c(g1,g2))))
```

### athi$cohensW - effect size for contingency tables

Effect sizes measure for categorical data.

Calculate the effect size for 2x2 and larger contingency tables as well as for single variables.

The function `athi$cohensW` (omega) calculates the effect size for contingency tables. 
Due to Cohen's rule of thumb values of around 0.1 are considered to stand 
for small effects, values of around 0.3 represent medium effects and values 
above 0.5 or higher represent large effects. 

Please note that these rules of thumb are not useful for highly dependent outcome 
variables (death for instance) these rules might not be useful and as well lower
values might be of practical relevance.

Usage: *athi_cohensW(x,p=NULL)*

```{r}
data(Titanic)
### first class data only
Titanic[1,1,,]
athi$cohensW(Titanic[1,1,,])
### Data from New Eng. J. Med. 329:297-303, 1993
azt=as.table(matrix(c(76,399,129,332), byrow=TRUE,ncol=2))
rownames(azt)=c("AZT","Placebo")
colnames(azt)=c("DiseaseProgress", "NoDiseaseProgress")
azt
athi$cohensW(azt)
### number of boys (25) and girls (15) in a hospital which deviates 
### from 50/50 or 51/49 ratios
prop.table(c(25,15))
(prop.table(c(25,15))-0.5)*2
athi$cohensW(c(25,15),p=0.5)
athi$cohensW(c(25,15),p=c(0.51,0.49))
### most extrem case 40 boys and 0 girls
athi$cohensW(c(40,0),p=0.5)
athi$cohensW(c(40,0),p=c(0.51,0.49)) # max value here around 2*0.49
```

### athi$corr - matrix of pairwise correlations

Pairwise  correlations for a matrix or data frame with estimate,  p-values and
confidence intervals.

Method to determine pairwise correlations and their statistical properties.
This function is an extension to the `cor` function as it as well calculates all p-values and the confidence intervals for the
correlation coefficient. 

The confidence  interval for the Spearman  correlation is calculated using the
Fisher transformation, see [https://en.wikipedia.org/wiki/Fisher_transformation](https://en.wikipedia.org/wiki/Fisher_transformation).

Usage: *athi_corr(x,method='pearson',use='pairwise.complete.obs')*

```{r}
options(width=90)
data(swiss)
res=athi$corr(swiss)
ls(res)
lapply(res[1:4],round,2)
```

### athi$cor_plot - improved xy-plot with effect size

visualize a correlation with abline and main statistics.

This function is plotting the standard xy-plot for two numerical variables and adds on top the 
the main statistics, like the r-value, the confidence interval and the significance level.

Usage: *athi_cor_plot(x,y,method="pearson",col='blue',grid=TRUE,main=NULL,pch=19,...)*

```{r fig.width=6,fig.height=4}
data(swiss)
athi$cor_plot(swiss$Fertility,swiss$Agriculture,
     xlab="Fertility",ylab="Agriculture")
```

### athi$corr_plot - plot of correlation matrices

Visualize a correlation matrix

This function is plotting the pairwise correlations for the given
symmetric correlation matrix.

Usage: *athi_corr_plot(mt,text.lower=TRUE, text.upper=FALSE, pch=19,p.mat=NULL,alpha=0.05, cex.sym=5,cex.r=1,cex.lab=1.4,...)*

```{r fig=TRUE,fig.width=6,fig.height=4}
par(mai=rep(0.2,4))
data(swiss)
sw=swiss
colnames(sw)=abbreviate(colnames(swiss),6)
options(warn=-1) # avoid spearman warnings
cr=athi$corr(sw,method='spearman')
athi$corr_plot(cr$estimate,cex.sym=4,text.lower=TRUE,
   cex.r=0.9,cex.lab=0.8,p.mat=cr$p.value)
options(warn=0)
```

### athi$cv - coefficient of variation

Calculate the coefficient of variation.

Usage: *athi_cv(x,na.rm=FALSE)*

```{r}
cv=athi$cv
cv(rnorm(20,mean=100,sd=4))
cv(c(1,2,3,4))
cv(c(1,2,3,4,NA))
cv(c(1,2,3,4,NA),na.rm=TRUE)
```

### athi$df2md - creating Markdown tables

Convert a data frame or a matrix into a Markdown table.

This function can be used within Rmarkdown documents to display easily
a simple Markdown table. For more advanced use cases you should other commands
such as the kable method from the knitr package.

Usage: *athi_df2md(x,caption='',rownames=TRUE)*

```{r}
data(swiss)
athi$df2md(head(swiss),caption="<center>**Table: Sample swiss data**</center>")
```

Embedded with code chunk options `echo=FALSE,results="asis" in this document it looks like this:

```{r echo=FALSE,results="asis"}
data(swiss)
athi$df2md(head(swiss),caption="<center>**Table: Sample swiss data**</center>")
```

### athi$epsilon_squared - effect size for Kruskal test

Effect sizes measure for Kruskal test.

Calculate the effect size epsilon-squared for variables of a kruskal.test
The function `athi$epsilon_squared` calculates the effect size for a Kruskal test.
Cohen's rule of thumb for interpretation is: around 0.01 small, around 0.09 medium and around 0.25 or higher we have a large effect.
You can convert Epsilon-squared to a Pearson r coefficient by using the sqrt of The Epsilon-square value.

Please note that these rules of thumb are not useful for highly dependent outcome 
variables (death for instance) these rules might not be useful and as well lower
values might be of practical relevance.

Usage: *athi_epsilon_squared(x,y)*

```{r}
data(iris)
athi$epsilon_squared(iris$Sepal.Length,iris$Species)
# two factor example as well for wilcox.test possible
data(ToothGrowth)
athi$epsilon_squared(ToothGrowth$len,as.factor(ToothGrowth$dose))
# close to r-square of spearman!
cor(ToothGrowth$len,ToothGrowth$dose,method="spearman")^2
```


### athi$eta_squared - effect size for ANOVA

Effect sizes measure for ANOVA.

Calculate the effect size for an ANOVA or a linear model with two variables.

The function `athi$eta_squared` (omega) calculates the effect size for an ANOVA.
Cohen's rule of thumb for interpretation is: around 0.01 small, around 0.09 medium and around 0.25 or higher we have a large effect.
You can convert Eta-squared to a Pearson r coefficient by using the sqrt of eta-square.

Please note that these rules of thumb are not useful for highly dependent outcome 
variables (death for instance) these rules might not be useful and as well lower
values might be of practical relevance.

Usage: *athi_eta_squared(x,y=NULL)*
 
```{r}
data(iris)
etaSquared=athi$eta_squared
etaSquared(iris$Sepal.Length,iris$Species)
etaSquared(lm(iris$Sepal.Length ~ iris$Species))
summary(lm(iris$Sepal.Length ~ iris$Species))
etaSquared(aov(iris$Sepal.Length ~ iris$Species))
etaSquared(aov(Sepal.Length ~ Species+Sepal.Width+Petal.Length,data=iris))
```

### athi$fmt - formatting strings

Python like formatting of strings using curly braces

This is a functions for formatted output avoiding a lot of quotes
during creation of strings based on variables. It mimics the Python format command.

Usage: *athi_fmt(x,...)*

```{r}
athi$fmt('I can say {} {}!',"Hello", "World")
athi$fmt('I can say {2} {1}!',"World", "Hello")
athi$fmt("pi is '{}'!",sprintf("%3.5f",pi))
```

### athi$impute - impute missing values

Replaces missing values with a reasonable guess by different imputation methods.

This method aids in replacing missing values with a reasonable guess by different imputation methods such as 
the simple and not recommended methods mean and median, where NA's are replaced with the 
mean or median for this variable or the more recommended methods using rpart decision trees
or knn using a correlation distance based k-nearest neighbor approach. 
The rpart method can be as well used to replace missing values for categorical variables.
In case of median and mean imputations for categorical variables the modus is used, 
so missing values are replaced with the most often category. This is rarely reasonable.

Usage: *athi_impute(x,method="rpart",k=5,cor.method="spearman")*

```{r}
data(iris)
ir=athi$introNAs(iris[,1:4],prop=0.1)
summary(ir)
ir=athi$impute(ir)
summary(ir)
ir=iris
ir[1,3]=NA; ir[2,4]=NA; ir[c(1,3),5]=NA
head(ir)
head(athi$impute(ir,method="rpart"))
```


### athi$input - read input from user in scripts

Reads input strings from the user, even in non-interactive mode.

This  function  should be used  instead of  readline in files which  should be
started with Rscript to get inout from the user.

Usage: *athi$input(prompt)*

```{r eval=FALSE}
### the code below is not evaluated
x = athi$input("Enter a number: ")
x = as.numeric(x)
print(x*x)
```

### athi$introNAs - introduce some NAs into data

Introduce missing values into data frames or matrices.

Introduces  NA's  into  the  given  data  frame  or  matrix  with a  specified
proportion for evaluation of imputing methods.

Usage: *athi_introNAs(x,prop=0.05)*

```{r}
data(iris)
ir=athi$introNAs(iris[,1:4],prop=0.1)
summary(ir)
dim(ir)
apply(ir,2,function (x) { length(which(is.na(x))) })
```


### athi$kroki - create diagrams, flowcharts

Create diagrams using the online tool [kroki](https://kroki.io).

This function is creates a URL which can be easily embedded into Markdown code for displaying
diagrams supported by the online tool [kroki.io](https://kroki.io).
There is as well an online diagram editor, see here [niolesk](https://niolesk.top/).

```{r label=kroki}
 url1=athi$kroki('
 digraph g { 
    rankdir="LR";
    node[style=filled,fillcolor=beige];
    A -> B -> C ; 
 }',
 type="graphviz")
 url2=athi$kroki("
 +---------+    +--------+
 |    AcEEF+--->+   BcFEE+
 +---------+    +--------+
 ")
```

 To embed the image you can use Markdown code like here:

```
   # remove space before r letter
   ![ ](` r url1 `)
```

Here the output:
 
![ ](`r url1`)
 
And here the image for the second diagram, a Ditaa diagram:
 
![ ](`r url2`)
 
The diagram code can be read as well from a file here a Ditaa file:
 
```{r}
  cat('/---------\\  +--------+   +--------+
|  cFDD   |  |        |   | {s}    |
|  Hello  +->| World! +-->|   DB   |
|         |  | cDDF   |   |  cFFD  |  
\\---------/  +--------+   +--------+

',file="hw.ditaa")
url3=athi$kroki(filename="hw.ditaa")
nchar(url3)
```

![](`r url3`)

For images stored in files it can be even easier without intermediate variables like this:

```
  ![ ](`r athi$kroki(filename="hw.ditaa")`)
```

Images can be however as well directly plotted if the png library is available:
 
```{r fig=TRUE,fig.width=4,fig.height=4,out.width=200,fig.cap=""}
athi$kroki("digraph { plot -> kroki; }",type="graphviz",plot=TRUE)
```
 
You can as well use the PlantUML server, here an example:

```{r}
nocache.url=athi$kroki("
@startuml
Bob -> Alice : hello
@enduml
",server='plantuml',cache=FALSE,ext='svg')
print(nocache.url)
url=athi$kroki("
@startuml
Bob -> Alice : hello
@enduml
",server='plantuml',ext='svg')
url
```

![](`r url`)

See also:

- [ditaa documentation I](http://ditaa.sourceforge.net/)
- [ditaa documentation II](https://github.com/stathissideris/ditaa)
- [plantuml documention](https://plantuml.com/)
- [graphviz documentation](https://graphviz.org/documentation/)
- [kroki documentation](https://kroki.io/)
- [niolesk editor](https://niolesk.top/#https://kroki.io/plantuml/svg/eNplj0FvwjAMhe_5FVZP40CgaNMuUGkcdttp3Kc0NSVq4lRxGNKm_fe1HULuuD37-bOfuXPUm2QChEjRnlIMCDmdUfHNSYY6xh42a9Fsegflk-yYlOLlcHK2I2SGtX4WZm9sZ1o8uOzxxbuWAlIGj8cshs6M1jDuY2owyU2P8jAezdnn10j53X0hlBsZFW021Pq7HaVSNw-KN-OogG8F8BAGqT8dXhZjxW4cyJEW6kcC-yHWFagHqW0MfaThhYmaVyE26P_x27qaDmXeruqqAMMw1h-ZlRI4aF3dX7hOwm5XzfIKDctlNcshPT1tFa8JPYAj-Zf5F065sqM=)

### athi$lm_plot - linear model plot with CI

Plot a linear model with confidence intervals.

This is a convinience method to plot a linear model for confidence intervals for
the slope and for the predictions based on the model.

Usage: *athi_lm_plot(x,y=NULL, data=NULL,col="blue",pch=19,col.lm="red",col.plm="red",col.pi="blue",
                       grid=TRUE,polygon=TRUE,col.polygon="#cccccc33",xlab=NULL,ylab=NULL,...)*

```{r label=lmplot,fig.width=6,fig.height=3}
par(mfrow=c(1,2),mai=c(0.8,0.8,0.2,0.1))
data(iris) 
athi$lm_plot(iris$Sepal.Width, iris$Sepal.Length,
   col=as.numeric(iris$Species)+1,col.pi="bisque4",
   col.lm="black",xlab="Sepal.Width",ylab="Sepal.Length")
props=c(0.0441,0.0133,0.0183,0.0238,0.0389,
        0.0648,0.0275,0.0704,0.0796,0.036,0.0132,
        0.108,0.136,0.0383,0.1008)
years=2005:2019
athi$lm_plot(years,props,ylim=c(0,0.3),xlab="Year",ylab="Proportion",
   col.pi=NULL,col.plm='#FFB0B066',col.polygon='#FFB0B066')
```

### athi$mi - mutual information

Return the mutual information for two vectors, a matrix or data frame,  or a binned table.

Usage: `athi_mi(x,y=NULL,breaks=4,norm=FALSE)`{.r}

```{r label=mi}
rn1=rnorm(100,mean=10,sd=1);
rn2=rn1+0.5*rnorm(100)
cor(rn1,rn2) # high
cor(rn1,sample(rn2)) #random 
athi$mi(rn1,rn2) # high 
athi$mi(rn1,sample(rn2)) #random
athi$mi(rn1,rn2,breaks=4)
athi$mi(rn1,rn2,breaks=7)
data(swiss)
round(athi$mi(swiss),2)
round(athi$mi(swiss,norm=TRUE),2)
```

### athi$norm - data normalization

Normalizes given data frame or matrix using the choosen method.

Perform data  normalization  on the current data sets such as z-score, feature
scaling, quantile and so on.

Usage: *athi_norm(x,method="z",ties.method="average")*

```{r fig.width=6,fig.height=3}
par(mai=c(0.7,0.7,0.2,0.1))
data(iris)
ir=athi$norm(iris[,1:4],method="uv")
apply(ir,2,sd)
summary(ir)
ir2=athi$norm(iris[,1:4],method="fs")
boxplot(ir2)
summary(athi$norm(iris[,1:4],method="q"))
# NA's are handled as well
ir=iris[,1:4]
ir[2,3]=NA
ir[1,4]=NA
ir[3,2]=NA
head(ir)
head(athi$norm(ir,method="q"))
summary(athi$norm(ir,method="q"))
sdata=read.table(text="4 3 6 4 7
8 1 10 5 11
6 2 7 8 8
9 4 12 9 12
7 5 9 6 10
")
athi$norm(sdata,method="mp")
```

### athi$pca_biplot - PCA biplot
 
Improved biplot for pca objects.

The function `athi$pca_biplot` provides an improved biplot for
visualizing the pairwise scores of individual principal components of 
an object created using the function `prcomp`. In contrast to the default 
biplot function  this plot visualizes the data as points and not row numbers,
it allows to display groups using color codes and distribution ellipses.

Arguments:

```{r}
args(athi$pca_biplot)
```

Examples:

```{r label=biplot,fig=TRUE,fig.height=6,fig.width=12,fig.cap=""}
par(mai=c(0.8,0.8,0.2,0.6),mfrow=c(1,2))
data(iris)
pci=prcomp(iris[,1:4],scale=TRUE)
athi$pca_biplot(pci,col=rep(2:4,each=50),ellipse=TRUE,ell.fill=TRUE,
    arrow.fac=2.3,arrows=TRUE,main="biplot")
legend('topright',pch=19,col=2:4,levels(iris$Species))
# standard score plot
athi$pca_biplot(pci,col=rep(2:4,each=50),ellipse=FALSE,
   arrow.fac=2.3,arrows=FALSE,main="scoreplot")
```

### athi$pca_oncor - PCA on a correlation matrix 
 
Perform a PCA on a correlation matrix.

The function `athi$pca_oncor` does a PCA using eigenvector eigenvalue decomposition
on a correlation matrix. PCA is usually performed on Pearson correlation internally what
leads to a highly outlier sensitive analysis. If the user decides
to use a method like Spearman correlation or even bi-seriell, polychoric or for nominal data
effect size measures like Cohen's W this method here can be used. Note that this does not return new coordinates for the sample as the sample contribution is lost in the
correlation matrix. The method might however be used to check if the results between Pearson and Spearman PCA are similar or does 
outliers lead to a completly different result.

Arguments:
```{r}
args(athi$pca_oncor)
```

Examples:

```{r label=oncor}
data(USArrests)
C=cor(USArrests)
athi$pca_oncor(C)
D=cor(USArrests,method="spearman") 
athi$pca_oncor(D)
```
 
### athi$pca_pairs - pairs plot for PCA 
 
Improved pairs plot for PCA objects.
 
The function `athi$pca_pairs` provides an improved pairs plot for
visualizing the pairwise scores of the individual components of an analyses 
using the function `prcomp`. In contrast to the default pairs function 
this plot visualizes in the diagonal as well the variances and 
a density line for the component scores.


Arguments:

```{r}
args(athi$pca_pairs)
```
 
Examples:

```{r label=pca.pairs,fig=TRUE,fig.cap=""}
data(iris)
pci=prcomp(iris[,1:4],scale=TRUE)
athi$pca_pairs(pci,pch=15,groups=iris[,5],
   legend=TRUE,oma=c(5,4,4,4),col=as.numeric(iris[,5])+1)
```

### athi$pca_plot - improved screeplot for PCA
 
Improved bar or screeplot for pca objects.

The function `athi$pca_plot` provides an improved bar- or screeplot for
visualizing the variances of the individual components of an analyses 
using the function `prcomp`. In contrast to the default plot function 
this plot visualize cumulative and individual variances in percent.

Arguments:

```{r}
args(athi$pca_plot)
```

Examples:

```{r label=pca.plot,fig=TRUE,fig.width=11,fig.height=6,fig.cap=""}
data(iris)
par(mfrow=c(1,2))
pcai=prcomp(iris[,1:4],scale=TRUE)
athi$pca_plot(pcai)
athi$pca_plot(pcai,type="scree",legend=FALSE)
```

### athi$qr_plot - quantile regression plot

Plot quantile regression models.

This is a convinience method to plot quantile regression models and giving
optional percentile intervals for a given range of predictions.
the slope and the intercept for the different quantiles are returned as well if requested.

Usage: `athi_qr_plot(x,data,quantiles=c(0.05,0.1,0.5,0.9,0.95),pred=NULL,plot=TRUE,...)`

```{r label=qr-plot,fig.width=5,fig.height=3}
par(mai=c(0.8,0.8,0.3,0.1))
data(iris) 
athi$qr_plot(Sepal.Width ~ Sepal.Length,data=iris[51:151,])
res=athi$qr_plot(Sepal.Width ~ Sepal.Length,data=iris[51:151,],pred=c(5,5.5,6,6.5,7,7.5,8))
res$centiles
res$coef
```

### athi$randomize - randomize data column wise

Randomize data frame or matrix columns

This function can be used to randomize the data within the same column.

Usage: *athi_randomize(x)*

```{r}
data(iris)
round(cor(iris[,1:4]),2)
round(cor(athi$randomize(iris[,1:4])),2)
```

### athi$ref_score - age independent WHO reference scores

Age corrected z-scores based on WHO references.

Function to retrieve age corrected z-scores based on WHO references

Usage: *athi_ref_score(x,age,sex,type)*

```{r}
athi$ref_score(100,age=4,sex="M",type="height")
athi$ref_score(100,age=4,sex="F",type="height")
athi$ref_score(c(100,100,100),age=c(4,4.2,4),sex=c("F","F","M"),type="height")
head(athi$ref_table(sex="F",type="height"))
```

### athi$ref_table - WHO reference tables for childrens growth

Function to retrieve WHO reference tables.

Usage: *athi_ref_table(sex,type)*

```{r}
athi$ref_table(sex="M",type="height")
head(athi$ref_table(sex="F",type="height"))
```

### athi$report_pvalue - report p-values consistently

Function for reporting a p-value either giving the three alpha thresholds, 
<0.05, <0.01, or <0.001 or using the star syntax. 

Usage: *athi$report_pvalue(x,star=FALSE)*

```{r}
athi$report_pvalue(1/10000)
athi$report_pvalue(1/10000,star=TRUE)
athi$report_pvalue(0.02,star=TRUE)
athi$report_pvalue(0.12,star=TRUE)
athi$report_pvalue(c(0.001,0.01,0.3,0.02))
```

### athi$sem - standard error of the mean

Calculate the standard error of the mean for a given numerical vector.

Usage: *athi$sem(x,na.rm=FALSE)*
' 
```{r label=sem}
sem=athi$sem
sem(rnorm(50,mean=10,sd=3))
sem(rnorm(1000,mean=10,sd=3))
```

### athi$smartbind(x,y) - combine two data frames, matrices

Combine two data frames via rbind even with different column names.

This function combines two data frames or matrices even if they have different
column names. In each data frame or matrix missing columns are first filled up
with NA's. Thereafter data is fused using rbind. Column order is determined based 
on the first given data.

```{r label=smartbind}
ir1=cbind(rn1=rnorm(nrow(iris)),iris)
ir2=cbind(iris,rn2=rnorm(nrow(iris),mean=10))
head(athi$smartbind(ir1,ir2))
tail(athi$smartbind(ir1,ir2))
```

### athi$textplot - table plot for data

Write the data for a data frame or matrix into a plot.

This function can be used as a workaround to display data for small
data frames or matrices with let's say 3-8 rows or columns into a plot. 

Usage:  *athi$textplot(x,cex=1,caption=NULL,...)*

```{r label=textplot,fig=TRUE,fig.width=5,fig.height=3,fig.cap=""}
par(mai=rep(0.1,4))
athi$textplot(head(swiss),cex=0.7,caption="Table 1: Swiss data first six lines")
```

### athi$untab(x) - table to data frame

Convert a contingency table to a data frame one item per row.

This function takes a contingency table and expands it two a data frame
where every level combinations is created according to the number of entries in the
contingency table. You can reverse the process using the table command. The procedure is useful for
instance for performing sample bootstrapping.

Usage: *athi$untab(x)*

```{r label=untab}
M=matrix(c(4,3,1,2),ncol=2,dimnames=list(
        size=c('small','tall'),
        smoking=c('no','yes')))
M
N=athi$untab(M)
head(N)
summary(N)
table(N[,1],N[,2])
```

### athi$venn - Venn diagrams for sets

Plot Venn diagram for logical relations between two or three sets.

This function combines two data frames or matrices even if they have different
column names. In each data frame or matrix missing columns are first filled up
with NA's. Thereafter data is fused using rbind. Column order is determined based 
on the first given data.

Usage: *athi$venn(x,y=NULL,z=NULL,...)*

```{r label=venn,fig=TRUE,fig.width=8,fig.height=8,fig.cap=""}
par(mfrow=c(2,2),mai=rep(0.1,4))  
X=matrix(rnorm(2700),ncol=3)
colnames(X)=c("A","B","C")
X=X>0.4
Y=X[,1:2]
Y=Y>0.7
par(mfrow=c(2,2),mai=rep(0.1,4),pty='s')
athi$venn(X)
athi$venn(Y)
athi$venn(x=LETTERS[1:9],y=LETTERS[3:6],z=LETTERS[4:12],vars=c('Xvar','Yvar','Zvar'))
athi$venn(x=LETTERS[1:9],y=LETTERS[3:6],vars=c('Xvar','Yvar'))
```

###  athi$wilcoxR - effect size for Wilcoxon test
 
Calculate the effect size for a wilcox test (Rosenthal 1991).
Cohen's rule of thumb for interpretation is: approx 0.1 small, 0.3 medium and 0.5 and above is a large effect. 

Usage: *athi$wilcoxR(x,y=NULL,n=NULL)*

```{r label=wilcoxR}
set.seed(123)
rn1=rnorm(100,mean=10,sd=1)
rn2=rnorm(80,mean=11,sd=1)
rn2=c(rn2,rnorm(20,mean=12,sd=2)) # bimodal distribution
mean(rn1)
mean(rn2)
athi$wilcoxR(rn1,rn2) # two numerical vectors
wt=wilcox.test(rn1,rn2,exact=FALSE) 
athi$wilcoxR(wt,n=200) # a test object
cat = as.factor(c(rep('A',100),rep('B',100)))
athi$wilcoxR(x=c(rn1,rn2),y=cat) # num ~ cat
```

## EOF



