---
title: "Vignette for the functions of the athi package"
author: Detlef Groth, University of Potsdam, Germany
date: 2023-07-26
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
include-before: |
    <style> body { max-width: 1000px; font-family: Candara, sans-serif;} </style>
    <center>[athi functions](athi-functions.html) - [athi analysis](athi-analysis.html) </center>
    
---

```{r echo=FALSE,results="hide"}
library(knitr)
```

## Introduction

Below you find example code for the package for most of the  functions  which
should illustrate on how to use them.

You first should load the package using the standard  `library`  function. All
functions are provided  within an  environment  `athi` to see what methods are
available you can use the `ls` command:

```{r}
library(athi)
ls(athi)
```

Alternatively all functions have copies using the athi-underline  prefix which
you can as well view like this:

```{r}
ls(pattern="athi_")
```

The  advantage of the  environment  athi is, that you can easily extend it by
adding  your own data and  functions  to it. It is then  possible to save this
file  as RDS and to give it to  someone  else or as a  supplement  for a paper
without the need to create a full package.

Here an example:

```{r}
athi$hello = function () { print("Hello World") }
data(iris)
athi$data_iris = iris
ls(athi)
```

You      can       save       this       extended       environment       with
`saveRDS(athi,file="athi-extended.RDS")`  and you can  load  this  environment
then on other computer using the command `athi=readRDS("athi-extened.RDS")`


## Functions 

### athi$assoc_plot - contingency table plot with effect size measure

Create an association plot with Pearson residual coloring and legend.

This function updates the standard assocplot function from the graphics package 
with the ability to display residual colors. In blue and red are shown groups with 
residuals  above +4 or below -4 in light colors are shown residuals  between 2
and 4 for positive and -4 and -2 for negative  residuals.  At the bottom there
is the residual legend shown as well as the P-value for the Chisq-test and the
effect size measure Cohen's d.

Usage: *athi_assoc_plot(x,legend=TRUE,shade=TRUE,...)*

```{r fig=TRUE,fig.width=8,fig.height=6}
x <- margin.table(HairEyeColor, c(1, 2))
athi$assoc_plot(x)
```

### athi$bootstrap - resampling for empirical p-values or quantiles

Perform a resampling for the given data set and function. 

The function allows you to perform a resampling method without replacement to perform
a boostrap analysis for instance to cmpute a p-value or a confidence interval.

Usage: *athi_bootstrap(x,FUN=NULL,n=1000,...)*

```{r}
rn=rnorm(100,mean=10,sd=2)
t.test(rn)$conf.int
vals=athi$bootstrap(rn,FUN=mean)
summary(vals)
quantile(vals,c(0.025,0.975)) # 95% CI is very close
### confidence interval for spearman correlation
cor.test(swiss[,1],swiss[,2],method="spearman")
vals=athi$bootstrap(swiss[,c(1,2)],
      FUN=function(x) cor(x[,1],x[,2],method="spearman"))
summary(vals)
### insignificant as it crosses zero
quantile(vals,c(0.025,0.975)) # 95% CI shows insignifcant
```

### athi$box_plot - boxplot with effect size measure

Visualize a mean comparison with boxplot and main statistics.

This function is plotting the standard boxplot for a numerical data against a factor variabl and adds on top the 
the main statistics, like the Cohen's d, or the Eta-squared value, the significance level and for t-tests the  
confidence interval as well.

Usage: *athi_box_plot(x,y=NULL,data=NULL,col='grey80',grid=TRUE,main=NULL,...)*

```{r fig.width=6,fig.height=4}
data(iris)
athi$box_plot(Sepal.Length ~ Species, data=iris,col=2:4)
```

### athi$cdist - correlation distance measure

Correlation  distances  between  two  vectors or for all  variable  pairs in a
matrix.

Usage: *athi_cdist(x,method="pearson",type="abs")*

```{r}
### Euclidean distance
round(dist(t(scale(iris[,1:4]))),2)
### Correlation distance
round(athi$cdist(t(iris[,1:4])),2)
round(athi$cdist(t(iris[,1:4]),method="spearman",type="square"),2)
```

### athi$chr2ord - characters or factor to number conversion

Map characters or factors to numbers.

Create ordinal numerical variables out of character or factor variables based on a given mapping.

Usage: *athi_chr2ord(x,map)*

```{r}
status=c("never","rare","often","always")
x=sample(status,100,replace=TRUE)
x=c(NA,x,NA)
table(x,useNA='ifany')
map=c(never=0, rare=1, often=2,always=3)
table(athi$chr2ord(x,map),useNA='ifany')
```


### athi$cohensD - effect size for comparing two means

Effect sizes measure for difference between two means.

Calculate the effect size for the difference between two means, divided by the pooled standard deviation.

The function cohensD calculates the effect size for the difference between two means.
Due to Cohen's rule of thumb values of around 0.2 are considered to stand 
for small effects, values of around 0.5 represent medium effects and values of around 0.8 
and larger represent large effects. 

Please note that these rules of thumb are not useful for highly dependent 
outcome variables (death for instance) these rules might not be useful and as 
well lower values might be of practical relevance.

Usage: *athi_cohensD(x,g,paired=FALSE)*
 
```{r}
cohensD=athi$cohensD
set.seed(125)
data(sleep)
with(sleep,cohensD(extra,group))
x1=rnorm(100,mean=20,sd=1)
x2=rnorm(100,mean=22,sd=1)
g1=rep('A',100)
g2=rep('B',100)
### difference should be around 2SD
cohensD(c(x1,x2),as.factor(c(g1,g2)))
### biseriell correlation coefficient as alternative
### value is as well large
cor(c(x1,x2),as.numeric(as.factor(c(g1,g2))))
```

### athi$cohensW - effect size for contingency tables

Effect sizes measure for categorical data.

Calculate the effect size for 2x2 and larger contingency tables as well as for single variables.

The function `athi$cohensW` (omega) calculates the effect size for contingency tables. 
Due to Cohen's rule of thumb values of around 0.1 are considered to stand 
for small effects, values of around 0.3 represent medium effects and values 
above 0.5 or higher represent large effects. 

Please note that these rules of thumb are not useful for highly dependent outcome 
variables (death for instance) these rules might not be useful and as well lower
values might be of practical relevance.

Usage: *athi_cohensW(x,p=NULL)*

```{r}
data(Titanic)
### first class data only
Titanic[1,1,,]
athi$cohensW(Titanic[1,1,,])
### Data from New Eng. J. Med. 329:297-303, 1993
azt=as.table(matrix(c(76,399,129,332), byrow=TRUE,ncol=2))
rownames(azt)=c("AZT","Placebo")
colnames(azt)=c("DiseaseProgress", "NoDiseaseProgress")
azt
athi$cohensW(azt)
### number of boys (25) and girls (15) in a hospital which deviates 
### from 50/50 or 51/49 ratios
prop.table(c(25,15))
(prop.table(c(25,15))-0.5)*2
athi$cohensW(c(25,15),p=0.5)
athi$cohensW(c(25,15),p=c(0.51,0.49))
### most extrem case 40 boys and 0 girls
athi$cohensW(c(40,0),p=0.5)
athi$cohensW(c(40,0),p=c(0.51,0.49)) # max value here around 2*0.49
```

### athi$corr - matrix of pairwise correlations

Pairwise  correlations for a matrix or data frame with estimate,  p-values and
confidence intervals.

Method to determine pairwise correlations and their statistical properties.
This function is an extension to the `cor` function as it as well calculates all p-values and the confidence intervals for the
correlation coefficient. 

The confidence  interval for the Spearman  correlation is calculated using the
Fisher transformation, see [https://en.wikipedia.org/wiki/Fisher_transformation](https://en.wikipedia.org/wiki/Fisher_transformation).

Usage: *athi_corr(x,method='pearson',use='pairwise.complete.obs')*

```{r}
options(width=90)
data(swiss)
res=athi$corr(swiss)
ls(res)
lapply(res[1:4],round,2)
```

### athi$cor_plot - improved xy-plot with effect size

visualize a correlation with abline and main statistics.

This function is plotting the standard xy-plot for two numerical variables and adds on top the 
the main statistics, like the r-value, the confidence interval and the significance level.

Usage: *athi_cor_plot(x,y,method="pearson",col='blue',grid=TRUE,main=NULL,pch=19,...)*

```{r fig.width=6,fig.height=4}
data(swiss)
athi$cor_plot(swiss$Fertility,swiss$Agriculture,
     xlab="Fertility",ylab="Agriculture")
```

### athi$corr_plot - plot of correlation matrices

Visualize a correlation matrix

This function is plotting the pairwise correlations for the given
symmetric correlation matrix.

Usage: *athi_corr_plot(mt,text.lower=TRUE, text.upper=FALSE, pch=19,p.mat=NULL,alpha=0.05, cex.sym=5,cex.r=1,cex.lab=1.4,...)*

```{r fig=TRUE,fig.width=9,fig.height=6}
par(mai=rep(0.3,4))
data(swiss)
sw=swiss
colnames(sw)=abbreviate(colnames(swiss),6)
options(warn=-1) # avoid spearman warnings
cr=athi$corr(sw,method='spearman')
athi$corr_plot(cr$estimate,cex.sym=8,text.lower=TRUE,
   cex.r=1.5,p.mat=cr$p.value)
options(warn=0)
```

### athi$cv - coefficient of variation

Calculate the coefficient of variation.

Usage: *athi_cv(x,na.rm=FALSE)*

```{r}
cv=athi$cv
cv(rnorm(20,mean=100,sd=4))
cv(c(1,2,3,4))
cv(c(1,2,3,4,NA))
cv(c(1,2,3,4,NA),na.rm=TRUE)
```

### athi$df2md - creating Markdown tables

Convert a data frame or a matrix into a Markdown table.

This function can be used within Rmarkdown documents to display easily
a simple Markdown table. For more advanced use cases you should other commands
such as the kable method from the knitr package.

Usage: *athi_df2md(x,caption='',rownames=TRUE)*

```{r}
data(swiss)
athi$df2md(head(swiss),caption="<center>**Table: Sample swiss data**</center>")
```

Embedded with code chunk options `echo=FALSE,results="asis" in this document it looks like this:

```{r echo=FALSE,results="asis"}
data(swiss)
athi$df2md(head(swiss),caption="<center>**Table: Sample swiss data**</center>")
```

### athi$epsilon_squared - effect size for Kruskal test

Effect sizes measure for Kruskal test.

Calculate the effect size epsilon-squared for variables of a kruskal.test
The function `athi$epsilon_squared` calculates the effect size for a Kruskal test.
Cohen's rule of thumb for interpretation is: around 0.01 small, around 0.09 medium and around 0.25 or higher we have a large effect.
You can convert Epsilon-squared to a Pearson r coefficient by using the sqrt of The Epsilon-square value.

Please note that these rules of thumb are not useful for highly dependent outcome 
variables (death for instance) these rules might not be useful and as well lower
values might be of practical relevance.

Usage: *athi_epsilon_squared(x,y)*

```{r}
data(iris)
athi$epsilon_squared(iris$Sepal.Length,iris$Species)
# two factor example as well for wilcox.test possible
data(ToothGrowth)
athi$epsilon_squared(ToothGrowth$len,as.factor(ToothGrowth$dose))
# close to r-square of spearman!
cor(ToothGrowth$len,ToothGrowth$dose,method="spearman")^2
```


### athi$eta_squared - effect size for ANOVA

Effect sizes measure for ANOVA.

Calculate the effect size for an ANOVA or a linear model with two variables.

The function `athi$eta_squared` (omega) calculates the effect size for an ANOVA.
Cohen's rule of thumb for interpretation is: around 0.01 small, around 0.09 medium and around 0.25 or higher we have a large effect.
You can convert Eta-squared to a Pearson r coefficient by using the sqrt of eta-square.

Please note that these rules of thumb are not useful for highly dependent outcome 
variables (death for instance) these rules might not be useful and as well lower
values might be of practical relevance.

Usage: *athi_eta_squared(x,y=NULL)*
 
```{r}
data(iris)
etaSquared=athi$eta_squared
etaSquared(iris$Sepal.Length,iris$Species)
etaSquared(lm(iris$Sepal.Length ~ iris$Species))
summary(lm(iris$Sepal.Length ~ iris$Species))
etaSquared(aov(iris$Sepal.Length ~ iris$Species))
etaSquared(aov(Sepal.Length ~ Species+Sepal.Width+Petal.Length,data=iris))
```

### athi$fmt - formatting strings

Python like formatting of strings using curly braces

This is a functions for formatted output avoiding a lot of quotes
during creation of strings based on variables. It mimics the Python format command.

Usage: *athi_fmt(x,...)*

```{r}
athi$fmt('I can say {} {}!',"Hello", "World")
athi$fmt('I can say {2} {1}!',"World", "Hello")
athi$fmt("pi is '{}'!",sprintf("%3.5f",pi))
```

### athi$impute - impute missing values

Replaces missing values with a reasonable guess by different imputation methods.

This method aids in replacing missing values with a reasonable guess by different imputation methods such as 
the simple and not recommended methods mean and median, where NA's are replaced with the 
mean or median for this variable or the more recommended methods using rpart decision trees
or knn using a correlation distance based k-nearest neighbor approach. 
The rpart method can be as well used to replace missing values for categorical variables.
In case of median and mean imputations for categorical variables the modus is used, 
so missing values are replaced with the most often category. This is rarely reasonable.

Usage: *athi_impute(x,method="rpart",k=5,cor.method="spearman")*

```{r}
data(iris)
ir=athi$introNAs(iris[,1:4],prop=0.1)
summary(ir)
ir=athi$impute(ir)
summary(ir)
ir=iris
ir[1,3]=NA; ir[2,4]=NA; ir[c(1,3),5]=NA
head(ir)
head(athi$impute(ir,method="rpart"))
```


### athi$input - read input from user in scripts

Reads input strings from the user, even in non-interactive mode.

This  function  should be used  instead of  readline in files which  should be
started with Rscript to get inout from the user.

Usage: *athi$input(prompt)*

```{r eval=FALSE}
### the code below is not evaluated
x = athi$input("Enter a number: ")
x = as.numeric(x)
print(x*x)
```

### athi$introNAs - introduce some NAs into data

Introduce missing values into data frames or matrices.

Introduces  NA's  into  the  given  data  frame  or  matrix  with a  specified
proportion for evaluation of imputing methods.

Usage: *athi_introNAs(x,prop=0.05)*

```{r}
data(iris)
ir=athi$introNAs(iris[,1:4],prop=0.1)
summary(ir)
dim(ir)
apply(ir,2,function (x) { length(which(is.na(x))) })
```


### athi$kroki - create diagrams, flowcharts

Create diagrams using the online tool [kroki](https://kroki.io).

This function is creates a URL which can be easily embedded into Markdown code for displaying
diagrams supported by the online tool [kroki.io](https://kroki.io).
There is as well an online diagram editor, see here [niolesk](https://niolesk.top/).

```{r label=kroki}
 url1=athi$kroki('
 digraph g { 
    rankdir="LR";
    node[style=filled,fillcolor=beige];
    A -> B -> C ; 
 }',
 type="graphviz",server="kroki")
 url2=athi$kroki("
 +---------+    +--------+
 |    AcEEF+--->+   BcFEE+
 +---------+    +--------+
 ",server="kroki")
```

 To embed the image you can use Markdown code like here:

```
   # remove space before r letter
   ![ ](` r url1 `)
```

Here the output:
 
![ ](`r url1`)
 
And here the image for the second diagram, a Ditaa diagram:
 
![ ](`r url2`)
 
The diagram code can be read as well from a file here a Ditaa file:
 
```{r}
  cat('/---------\\  +--------+   +--------+
|  cFDD   |  |        |   | {s}    |
|  Hello  +->| World! +-->|   DB   |
|         |  | cDDF   |   |  cFFD  |  
\\---------/  +--------+   +--------+

',file="hw.ditaa")
url3=athi$kroki(filename="hw.ditaa",server="kroki")
nchar(url3)
```

![](`r url3`)

For images stored in files it can be even easier without intermediate variables like this:

```
  ![ ](`r athi$kroki(filename="hw.ditaa",server="kroki")`)
```

Images can be however as well directly plotted if the png library is available:
 
```{r fig=TRUE,fig.width=4,fig.height=4,out.width=200,fig.cap=""}
athi$kroki("digraph { plot -> kroki; }",type="graphviz",plot=TRUE,server="kroki")
```
 

See also:

- [ditaa documentation I](http://ditaa.sourceforge.net/)
- [ditaa documentation II](https://github.com/stathissideris/ditaa)
- [plantuml documention](https://plantuml.com/)
- [graphviz documentation](https://graphviz.org/documentation/)
- [kroki documentation](https://kroki.io/)
- [niolesk editor](https://niolesk.top/#https://kroki.io/plantuml/svg/eNplj0FvwjAMhe_5FVZP40CgaNMuUGkcdttp3Kc0NSVq4lRxGNKm_fe1HULuuD37-bOfuXPUm2QChEjRnlIMCDmdUfHNSYY6xh42a9Fsegflk-yYlOLlcHK2I2SGtX4WZm9sZ1o8uOzxxbuWAlIGj8cshs6M1jDuY2owyU2P8jAezdnn10j53X0hlBsZFW021Pq7HaVSNw-KN-OogG8F8BAGqT8dXhZjxW4cyJEW6kcC-yHWFagHqW0MfaThhYmaVyE26P_x27qaDmXeruqqAMMw1h-ZlRI4aF3dX7hOwm5XzfIKDctlNcshPT1tFa8JPYAj-Zf5F065sqM=)

### athi$lm_plot - linear model plot with CI

Plot a linear model with confidence intervals.

This is a convinience method to plot a linear model for confidence intervals for
the slope and for the predictions based on the model.

Usage: *athi_lm_plot(x,y=NULL, data=NULL,col="blue",pch=19,col.lm="red",col.plm="red",col.pi="blue",
                       grid=TRUE,polygon=TRUE,col.polygon="#cccccc33",xlab=NULL,ylab=NULL,...)*

```{r label=lmplot,fig.width=9,fig.height=6}
par(mfrow=c(1,2))
data(iris) 
athi$lm_plot(iris$Sepal.Width, iris$Sepal.Length,
   col=as.numeric(iris$Species)+1,col.pi="bisque4",
   col.lm="black",xlab="Sepal.Width",ylab="Sepal.Length")
props=c(0.0441,0.0133,0.0183,0.0238,0.0389,
        0.0648,0.0275,0.0704,0.0796,0.036,0.0132,
        0.108,0.136,0.0383,0.1008)
years=2005:2019
athi$lm_plot(years,props,ylim=c(0,0.3),xlab="Year",ylab="Proportion",
   col.pi=NULL,col.plm='#FFB0B066',col.polygon='#FFB0B066')
```

### athi$mi - mutual information

Return the mutual information for two vectors, a matrix or data frame,  or a binned table.

Usage: `athi_mi(x,y=NULL,breaks=4,norm=FALSE)`{.r}

```{r label=mi}
rn1=rnorm(100,mean=10,sd=1);
rn2=rn1+0.5*rnorm(100)
cor(rn1,rn2) # high
cor(rn1,sample(rn2)) #random 
athi$mi(rn1,rn2) # high 
athi$mi(rn1,sample(rn2)) #random
athi$mi(rn1,rn2,breaks=4)
athi$mi(rn1,rn2,breaks=7)
data(swiss)
round(athi$mi(swiss),2)
round(athi$mi(swiss,norm=TRUE),2)
```

### athi$norm - data normalization

Normalizes given data frame or matrix using the choosen method.

Perform data  normalization  on the current data sets such as z-score, feature
scaling, quantile and so on.

Usage: *athi_norm(x,method="z",ties.method="average")*

```{r fig.width=6,fig.height=4}
data(iris)
ir=athi$norm(iris[,1:4],method="uv")
apply(ir,2,sd)
summary(ir)
ir2=athi$norm(iris[,1:4],method="fs")
boxplot(ir2)
summary(athi$norm(iris[,1:4],method="q"))
# NA's are handled as well
ir=iris[,1:4]
ir[2,3]=NA
ir[1,4]=NA
ir[3,2]=NA
head(ir)
head(athi$norm(ir,method="q"))
summary(athi$norm(ir,method="q"))
sdata=read.table(text="4 3 6 4 7
8 1 10 5 11
6 2 7 8 8
9 4 12 9 12
7 5 9 6 10
")
athi$norm(sdata,method="mp")
```

### athi$qr_plot - quantile regression plot

Plot quantile regression models.

This is a convinience method to plot quantile regression models and giving
optional percentile intervals for a given range of predictions.
the slope and the intercept for the different quantiles are returned as well if requested.

Usage: `athi_qr_plot(x,data,quantiles=c(0.05,0.1,0.5,0.9,0.95),pred=NULL,plot=TRUE,...)`

```{r label=qr-plot,fig.width=9,fig.height=6}
data(iris) 
athi$qr_plot(Sepal.Width ~ Sepal.Length,data=iris[51:151,])
res=athi$qr_plot(Sepal.Width ~ Sepal.Length,data=iris[51:151,],pred=c(5,5.5,6,6.5,7,7.5,8))
res$centiles
res$coef
```

### athi$randomize - randomize data column wise

Randomize data frame or matrix columns

This function can be used to randomize the data within the same column.

Usage: *athi_randomize(x)*

```{r}
data(iris)
round(cor(iris[,1:4]),2)
round(cor(athi$randomize(iris[,1:4])),2)
```

### athi$ref_score - age independent WHO reference scores

Age corrected z-scores based on WHO references.

Function to retrieve age corrected z-scores based on WHO references

Usage: *athi_ref_score(x,age,sex,type)*

```{r}
athi$ref_score(100,age=4,sex="M",type="height")
athi$ref_score(100,age=4,sex="F",type="height")
athi$ref_score(c(100,100,100),age=c(4,4.2,4),sex=c("F","F","M"),type="height")
head(athi$ref_table(sex="F",type="height"))
```

### athi$ref_table - WHO reference tables for childrens growth

Function to retrieve WHO reference tables.

Usage: *athi_ref_table(sex,type)*

```{r}
athi$ref_table(sex="M",type="height")
head(athi$ref_table(sex="F",type="height"))
```

### athi$report_pvalue - report p-values consistently

Function for reporting a p-value either giving the three alpha thresholds, 
<0.05, <0.01, or <0.001 or using the star syntax. 

Usage: *athi_report_pvalue(x,star=FALSE)*

```{r}
athi$report_pvalue(1/10000)
athi$report_pvalue(1/10000,star=TRUE)
athi$report_pvalue(0.02,star=TRUE)
athi$report_pvalue(0.12,star=TRUE)
athi$report_pvalue(c(0.001,0.01,0.3,0.02))
```

## EOF



